NOPT048 Notes

### 2020-04-19 misc/study-notes/math457.tex
Added new sections

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Id{\mathop {\rm Id}}
\def\Sym{\mathop {\rm Sym}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}

@@

\floattext5 \noindent {\ninebf Note.} \ninept These notes are quite rough and skip over a lot of details. \ninept Most proofs are either omitted or distilled to their main ideas.

@@

\section 5. Modules over PIDs

@@

>>>>

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Id{\mathop {\rm Id}}
\def\Ev{\mathop {\rm Ev}}
\def\Sym{\mathop {\rm Sym}}
\def\Aut{\mathop {\rm Aut}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}
\def\F{{\bf F}}

@@

\floattext5 \noindent {\ninebf Note.} \ninept These notes are rough and may skip over some details. \ninept Some proofs are either omitted or distilled to their main ideas.

@@

\section 5. Modules Over PIDs

@@

\section 6. Fields and Polynomials

Because the kernel of a homomorphism is an ideal, then any nontrivial homomorphism $f:K\to R$ is injective when $K$ is a field. If $R = L$ is a field, then $K\subseteq L$ is a subfield and we call $L$ an {\it extension} of $K$. We will  often denote this by $L/K$. If $L/K$ is a field extension then $L$ is a vector space over $K$ and the dimension $[L:k] = \dim_K L$ is called the {\it degree} of the extension. Every there is a basis $(\alpha_i)$ of $L$ such that any element $l\in L$ can be expressed as $\lambda_1\alpha_1 + \cdots \lambda_n\alpha_n$ where $\lambda_i\in K$. If $K\subseteq L\subseteq M$ is a chain of extensions and $(\beta_j)$ is a basis of $M$ over $L$, then it can shown that $(\alpha_i\beta_j)$ is a basis of $M$ over $K$. So $[M:K] = [M:L][L:K]$.

A field is {\it prime} if it contains no proper nontrivial subfields. Any field $K$ is an extension of a prime field that contains $1$, $1+1$, etc. as well as their inverses. If the characteristic of $K$ is 0, then the prime field is $\Q$, and if the characteristic is a prime $p$, then the prime field is $\F_p$.

\proclaim Lemma F. Let $K$ be a field of characteristic $p$. The Frobenius map $x\mapsto x^p$ is a field homomorphism (so it is injective).

\proof For any $x,y\in K$, we have $(x+y)^p = x^p + y^p$ (by the binomial theorem) and $(xy)^p = x^py^p$.\slug

Let $L/K$ be an extension and $S\subseteq L$ be a set. Then $K(S)$ is the subfield of $L$ generated by $S$. The extension field is finitely generated if $S$ is finite. If $S$ consists of a single element $\alpha$, then $L = K(\alpha)$ is called a {\it simple} extension and $\alpha$ is a {\it primitive element}. For $\alpha_1,\ldots,\alpha_n$, the extensions $K(\alpha_1,\ldots,\alpha_n)$ and $K(\alpha_1)\cdots(\alpha_n)$ are the same and the order in which the elements are adjoined does not matter.

If $K$ is a field then $K[x]$ is a PID. So for any irreducible polynomial $f\in K[x]$, $(f)$ is maximal and $L = K[x]/(f)$ is a field extension. This is called the {\it Kronecker construction}.

\proclaim Lemma R. Every $f\in K[x]$ admits a root in a finite-degree extension.

\proof We may assume that $f$ is irreducible. It is of finite degree so $L = K[x]/(f)$ is a finite degree extension and if $\alpha = x \bmod f$, then $f(\alpha) = 0$ in $L$.\slug

Kronecker's construction is universal in the following sense. Let $L/K$ be an arbitrary extension and let $\alpha\in L$. Consider the map $\Ev_\alpha : K[\alpha] \to L$ that takes a polynomial $f$ to $f(\alpha)$. Since $K[x]$ is a PID, the $\ker(\Ev_\alpha)$ is principal and equals $(f_\alpha)$ for some polynomial $f_a\in K[x]$. Because $L$ is an integral domain, one of two things may happen. The first is that $f_\alpha$ is an irreducible polynomial, in which case we say that $\alpha$ is {\it algebraic}. The second is that the kernel is trivial and in this case we call $\alpha$ {\it transcendental}.

When $\alpha$ is algebraic, the unique monic irreducible polynomial $f_\alpha$ such that $f_\alpha(\alpha) = 0$ is called the {\it minimal polynomial} of $\alpha$ and $K(\alpha) \subseteq L$ is obtained by the Kronecker construction
$$K(\alpha)\cong K[x]/(f_\alpha).$$
If $\alpha$ is transcendental, $\Ev_\alpha : K[x] \inj L$ is injective and it extends to the fraction field $K(x) \inj L$ by taking $f/g \mapsto f(\alpha)/g(\alpha)$ (since $g(\alpha) \neq 0$ whenever $g\neq 0)$. Liouville established the existence of transcendental numbers in 1844 by proving that
$$L = \sum_{n\geq 0} {1\over 10^{n!}}$$
is transcendental. We have $\Q(L)\cong \Q(x) \subseteq \R$. Other famous transcendental numbers are $\pi$ and $e$.

An extension $L/K$ is {\it algebraic} if every $\alpha\in L$ is algebraic over $K$ and the following lemma proves some properties of algebraic extensions.

\proclaim Lemma A. Assume that an extension $L$ is generated by $\alpha_1,\ldots,\alpha_n$ over $K$. The following are equivalent:
\medskip
\item{i)} The elements $\alpha_1,\ldots,\alpha_n$ are algebraic over $K$.
\smallskip
\item{ii)} The degree $[L:K]$ is finite.
\smallskip
\item{iii)} Every $\alpha\in L$ is algebraic over $K$.
\medskip

\proof Suppose (i) holds. We have
$$K \subseteq K(\alpha_1)\subseteq K(\alpha_1, \alpha_2) \subseteq \cdots \subseteq L,$$
and since each $\alpha_i$ is algebraic over $K$, it is algebraic over $K(\alpha_1,\ldots,\alpha_{i-1}$, whence
$$\big[ K(\alpha_1,\ldots,\alpha_i) : K(\alpha,\ldots,\alpha_{i-1})\big] < \infty.$$
By the multiplicativity of the degree, $[L:K] \leq \infty$.

Suppose (iii) fails, i.e.\ some $\alpha\in L$ is not algebraic. Then $K(x) \cong K(\alpha) \inj L$ is an injection into $L$, contradicting the fact that $[K(x) : K] < \infty$. Thus (ii) implies (iii).

That (iii) implies (i) is obvious, so we are done.\slug

To construct extensions of a field, we need to find irreducible polynomials. Over $\Q$, we can consider $x^n - p$ where $p$ is a prime. Then $\Q(\root n\of p)$ is an extension of degree $n$ over $Q$. A general check for irreducibility is given by the following criterion.

\parenproclaim Theorem E (Eisenstein's criterion). Let $R$ be an integral domain and let $f\in R[x]$ be a monic polynomial of degree $n$. If there exists a prime ideal $\frak p$ such that $f = x^n \bmod \frak p$ and $f(0) \notin \frak p^2$, then $f$ is irreducible.

\proof Suppose, towards a contradiction, that $f = ab$ is reducible. Then, we have $x^n = \bar a \bar b$, where the bar indicates the polynomials modulo $\frak p$. In particular, $\bar a\bar b$ has zero constant term. The ideal $\frak p$ is prime, so $R/\frak p$ is an integral domain, so both $\bar a$ and $\bar b$ have zero constant term modulo $\frak p$, meaning that the constant terms of $a$ and $b$ both belong to $\frak p$. This is a contradiction, since it is clear that the constant term of $f$ belongs to $\frak p^2$.\slug

We can use Eisenstein's criterion to show that cyclotomic polynomials of the form
$$\Phi_p(x) = {x^p -1\over x-1} = x^{p-1} + x^{p-2} + \cdots + x + 1$$
for $p$ prime are irreducible. The criterion does not immediately apply, but if we consider
$$\Phi_p(x+1) = x^{p-1} + px^{p-2} + {p(p-1)\over 2}x + p,$$
we find that Eisenstein's criterion applies, so $\Phi_p(x+1)$ is irreducible and this implies that $\Phi_p(x)$ is irreducible, since any factorisation for $\Phi_p(x)$ would give a factorisation for $\Phi_p(x+1)$ (replacing $x$ with $x+1$).

There are other many other criteria for reducibility/irreducibility; we give two more famous ones.

\parenproclaim Theorem C (Cohn's criterion). Suppose that $p$ is prime and in some base $b$,
$$p = a_nb^n + \cdots + a_1b + a_0.$$
Then $f = a_nx^n + \cdots + a_1x + a_0$ in $\Z[x]$.\slug

\parenproclaim Lemma G (Gauss' lemma). Let $R$ be a UFD with fraction field $K$ and let $f\in R[x]$ be a polynomial of degree $n$ such that $\gcd(a_n, \ldots, a_0) = 1$. If $f$ is reducible in $K[x]$, then $f$ is reducible in $R[x]$.\slug

Gauss' lemma is often applied with $R = \Z$ and $K = \Q$.

\section 7. Splitting Fields

We begin with a lemma regarding the interchangeability of the roots of a polynomial.

\proclaim Lemma I. Let $f\in K[x]$ be a monic, irreducible polynomial and let $L/K$ be an extension of the field $K$. If $\alpha,\beta$ are two roots of $f$ in $L$, then there is a field isomorphism $K(\alpha) \cong K(\beta)$.

\proof This follows from the universality of the Kronecker construction: $K(\alpha) \cong K[x]/(f) \cong K(\beta)$.\slug

More generally, any field isomorphism $\phi : K\to L$ extends uniquely to a ring isomorphism $\bar\phi : K[x] \to L[x]$ defined by applying $\phi$ on the coefficients. Then $f\in K[x]$ is irreducible if and only if $\bar\phi(f)$ is irreducible. Let $\alpha$ be an arbitrary root of an irreducible polynomial $f\in K[x]$ and let $\beta$ be an arbitrary root of $\bar\phi(f)$. Then there exists a unique field isomorphism $\phi^* : K(\alpha) \to L(\beta)$ that takes $\alpha$ to $\beta$ and whose restriction to $K$ is $\phi$. A corollary of this fact is that if $f\in K[x]$ is irreducible, then all roots of $f$ have the same multiplicity in an algebraic closure (this will be expanded on later).

A {\it splitting field} for a polynomial $f\in K[x]$ is an extension $L/K$ such that
$$f = \prod_{i=1}^n (x - \alpha_i)$$
for $\alpha_i\in L$ and $L = K(\alpha_1,\ldots,\alpha_n)$.

\proclaim Proposition S. Every polynomial $f\in K[x]$ of degree $n$ admits a splitting field of degree at most $n!$.

\proof Let $\alpha_1$ be an abstract root of $f$ in $K(\alpha_1)$ obtained by the Kronecker construction. Then $f = (x-\alpha_1)f_1$ for some $f_1\in K(\alpha_1)$. Let $\alpha_2$ be a root of an irreducible factor of $f_1$ and extend the field to $K(\alpha_1,\alpha_2)$. This process happens at most $n$ times, by which time we will have found a splitting field $L$ of $f$. We have
$$K \subseteq K(\alpha_1) \subseteq K(\alpha_1,\alpha_2) \subseteq \cdots \subseteq L,$$
and the degree of each $f_i$ is $n-i$. So by multiplicity of the degrees we have $[L:K] \leq n!$.\slug

For example, the polynomial $f = x^4 - 1$ has roots $\pm 1, \pm i$. When $K = \Q$, the abstract bound for the degree of the splitting field is $4! = 24$, but in fact $\Q(i)$ is a splitting field for $f$, of degree 2. More generally, when $f = x^n - 1$, the roots are the $n$th roots of unity $1, \omega, \ldots, \omega^{n-1} \in\C$, where $\omega = e^{2ki\pi/n}$ for $i = 0,\ldots,n-1$. The roots form a group isomorphic to $\Z/n\Z = \langle\omega\rangle$ and $K(\omega)$ is the splitting field of $x^n - 1$, since if one root is added, all of its powers come along for the ride. If $n$ is prime, then the degree of this splitting field is $n-1$; in general, the degree is equal to $\varphi(n)$ where $\varphi$ denotes Euler's totient function.

The following theorem shows that splitting fields are unique up to $K$-isomorphism.

\parenproclaim Theorem K (Kronecker, {\rm 1887}). Let $f\in K[x]$ be an irreducible polynomial. Let $\alpha$ and $\alpha'$ be two roots of $f$ in two splitting fields $L/K$ and $L'/K$ respectively. Assume the existence of a map $\theta_0 \in \Aut(K)$ that fixes coefficients of $f$. Then there exists an isomorphism $\theta : L \to L'$ such that $\theta_{|K} = \theta_0$ and $\theta(\alpha) = \alpha'$.

\proof Extend $\theta_0$ to $\bar{\theta_0} : K[x]\to K[x]$. Then since $\theta_0$ fixes the coefficients of $f$, $\bar{\theta_0}(f) = f$. By the uniqueness of the Kronecker extension, there is an extension $\theta_1 : K(\alpha) \to K(\alpha)$ such that $\theta_1(\alpha) = \alpha'$ and $\theta_1_{|K} = \theta_0$. Write $f = (x - \alpha)f_1
